\section{Background and Problem Formulation}
\label{sec:background}

\subsection{Sampling-Based Uncertainty in RAG}
\label{sec:uq_background}

Let $q$ be a question, $c$ a retrieved context, and $r$ a model response.
Sampling-based uncertainty estimators operate on repeated draws from
$p(r \mid q)$.
In RAG, however, the response distribution is mediated by retrieval:
\begin{equation}
  p(r \mid q)
  = \sum_{c \in \mathcal{C}(q)} p(r \mid q, c)\, p(c \mid q),
  \label{eq:marginal}
\end{equation}
where $\mathcal{C}(q)$ is the finite set of candidate contexts the corpus and
retrieval policy make available for question $q$.
In words, the response distribution is a mixture over the contexts the
retriever might return, weighted by their retrieval probability.
Methods such as semantic entropy therefore reflect both decoder stochasticity
and whatever variability retrieval introduces.
In vanilla RAG this can be useful: different top-$k$ chunks yield different
answers when evidence is weak or unstable.

\subsection{Why KG-RAG Changes the Interpretation}
\label{sec:determinism}

KG-RAG changes how that disagreement should be read.
Vanilla RAG returns a ranked list of passages; KG-RAG can return a structured
state whose elements have names and types, encoding claims as triples
$(\mathrm{subject}, \mathrm{relation}, \mathrm{object})$.
Anchors, relation labels, triples, paths, and the passages that license them
are checkable intermediate objects, not merely more prompt text.
In the entity-first regime studied here, retrieval is mediated by entity
matching and graph expansion over a fixed graph $\mathcal{G}$.
When that process is highly stabilised, one question-conditioned context can
dominate:
\begin{equation}
  c^\star = \arg\max_{c \in \mathcal{C}(q)} p(c \mid q).
\end{equation}
As the residual mass $1 - p(c^\star \mid q)$ shrinks, the marginal in
Equation~\eqref{eq:marginal} is dominated by a single term,
\begin{equation}
  p(r \mid q) \approx p(r \mid q, c^\star),
  \label{eq:collapsed}
\end{equation}
and repeated samples mostly expose decoder variability under fixed evidence
(Appendix~\ref{sec:appendix_residual_bound} gives the corresponding
total-variation bound for the small-residual regime).
Equation~\eqref{eq:collapsed} is a stylised limit, not a claim about KG-RAG
as a category, and the experiments do not estimate $p(c \mid q)$ directly.
It formalises the regime of interest: once retrieval stabilises, low answer
variance can mean either good evidence or the same retrieval-side mistake on
every sample.

\subsection{A Three-Family Analysis Frame for KG-RAG}
\label{sec:decomposition}

\input{figures/decomposition}

``Uncertainty in KG-RAG'' is not a single quantity: entity anchoring, graph
expansion, evidence assembly, and generation can fail for different reasons.
The analysis tracks the three families against three observable
objects: answer, evidence, and retrieval state
(Table~\ref{tab:decomposition}).
The factorisation below is a bookkeeping device for these observable stages,
not a distinction implied by probability theory alone.
Let $s$ denote the graph retrieval state (anchors, paths, and subgraph), $c$
the textual evidence assembled from it, and $r$ the answer. The pipeline is
then
\begin{equation}
  p(r, c, s \mid q) = p(s \mid q)\, p(c \mid q, s)\, p(r \mid q, c, s).
\end{equation}
Answer-state uncertainty primarily probes variability in $p(r \mid q, c, s)$;
evidence-state measures probe whether the produced evidence $c$ is locally
consistent with the answer $r$; retrieval-state measures probe whether the
retrieval state $s$ supplies coherent graph support in the first place.
These are consistency diagnostics over the retrieved state, not direct
correctness tests, and the split is not a claim of independence: evidence-state
signals depend on retrieval, retrieval-state scores depend on entity extraction
and graph coverage, and answer variation depends on retrieval stability.
It is a reporting partition, not a causal ontology; a system may produce
low-variance answers from the wrong subgraph, retrieve a plausible path whose
passages do not entail the answer, or retrieve good evidence while the answer
remains unstable.
