\section{Uncertainty Measures}
\label{sec:metrics}

Eight diagnostic measures are evaluated across three observable objects in a
QA episode: sampled answers, retrieved graph state, and retrieved evidence
(Table~\ref{tab:metric_taxonomy}).
They are not eight independent estimates of one hidden confidence variable.
Within-family dependencies are explicit: the P(True)-style proxy is monotone
in DSE under black-box sampling, and VN-Entropy and SD-UQ are both geometric
statistics of the same response matrix.
The headline tables report family representatives rather than treating all
eight rows as separate evidence.
The answer-state scores mostly come from standalone LLM uncertainty
\citep{kuhn2023semantic,manakul2023selfcheckgpt,kadavath2022language,moskvoretskii2025adaptive};
the retrieval- and evidence-state scores ask whether the generated answer is
reachable in the graph and supported by the retrieved text.
Let $\bm{r} = \{r_1,\ldots,r_N\}$ denote $N$ responses sampled from the
LLM for a question $q$ with retrieved context $c$.
Let $\bm{v}_i \in \mathbb{R}^d$ denote the $\ell_2$-normalised embedding of
$r_i$, and $\bm{V} = [\bm{v}_1\mid\cdots\mid\bm{v}_N]^\top \in \mathbb{R}^{N\times d}$
the stacked response embedding matrix.
All estimators return a non-negative scalar with higher values indicating
greater uncertainty.
The retrieval-state diagnostic GPS and the evidence-state estimator SEU are
bounded in $[0,1]$; DSE and VN-Entropy have maximum $\log N$; SRE-UQ and
SD-UQ are non-negative and unbounded.
Because the experiments use only $N=5$ samples, the answer-state estimators are
low-sample operational diagnostics, not high-precision distribution estimates.

\begin{table*}[t]
\centering
\small
\caption{%
  \textbf{Taxonomy of the eight uncertainty measures.}
  The first six rows are answer-state estimators; the last two instantiate
  retrieval-state and evidence-state diagnostics.
  Measured per-question costs are reported in Appendix
  Table~\ref{tab:compute_cost}.
  $\dagger$: metric or KG-RAG operationalisation defined here.
  $\ddagger$: prior estimator first benchmarked here for KG-RAG.%
}
\label{tab:metric_taxonomy}
\begin{tabular}{@{}llp{5.2cm}p{2.9cm}p{2.7cm}@{}}
\toprule
\textbf{Family} & \textbf{Measure} & \textbf{Formal sketch} & \textbf{Required inputs} & \textbf{Cost at inference} \\
\midrule
Entropy
  & DSE
  & $-\sum_k \hat{p}_k \log \hat{p}_k$, with $\hat{p}_k = |C_k|/N$
  & sampled answers, NLI
  & $N$ samples; $O(N^2)$ NLI calls \\
\midrule
Calibration
  & $\mathrm{P(True)}$
  & $1 - |\{i:\mathrm{cl}(r_i)=\mathrm{cl}(r_1)\}|/N$; black-box P(True)-style proxy
  & sampled answers, NLI
  & $N$ samples; shares DSE clustering \\
\midrule
Similarity
  & SelfCheckGPT
  & $\frac{1}{|\mathcal{P}|}\sum_{(i,j)\in\mathcal{P}}
      [\mathrm{NLI}(r_i,r_j)=\textsc{contradiction}]$
  & sampled answers, NLI
  & $N$ samples; $O(N^2)$ NLI calls \\
\midrule
Perturbation
  & SRE-UQ
  & $\frac{1}{M}\sum_{i\in\mathcal{T}}|\Delta_i|$; perturbation sensitivity of the kernel mean embedding
  & sampled answer embeddings
  & $N$ samples; embeddings only \\
\midrule
\multirow{2}{*}{Geometric}
  & VN-Entropy$^{\ddagger}$
  & $-\operatorname{tr}(\bm{\rho}\log\bm{\rho})$, with $\bm{\rho}=\bm{V}\bm{V}^\top/N$
  & sampled answer embeddings
  & $N$ samples; embeddings only \\
  & SD-UQ$^\dagger$
  & $\exp\!\bigl(\frac{1}{k}\sum_{i=1}^k \log(\eta_i+\varepsilon)\bigr)$ on question-orthogonal residuals
  & question embedding, sampled answer embeddings
  & $N$ samples; embeddings only \\
\midrule
Retrieval-state
  & GPS$^\dagger$
  & $\displaystyle 1 -
     \frac{\sum_{e:\mathrm{reach}_e} w_e \gamma^{|L_e-\hat L(q)|}}
          {\sum_e w_e}$; soft-linked, depth-matched answer-entity support
  & retrieved KG, entity embeddings, linked question/answer entities
  & 1 sample; per-entity graph queries, no LLM calls \\
\midrule
Evidence-state
  & SEU$^\dagger$
  & $\frac{1 - (n_E-n_C)/K}{2}$; entailment--contradiction deficit over retrieved chunks
  & retrieved chunks, answer, NLI
  & 1 sample; $K$ NLI calls \\
\bottomrule
\end{tabular}
\end{table*}

% =============================================================================
\subsection{Answer-State Uncertainty Estimators}
\label{sec:gen_metrics}

\paragraph{Semantic entropy and DSE.}
\citet{kuhn2023semantic} cluster sampled responses into meaning-equivalent
sets $\mathcal{C}=\{C_1,\ldots,C_K\}$ using bidirectional NLI entailment.
The API setting exposes no reliable per-sample likelihoods, so the reported
score is the count-weighted black-box variant,
\begin{equation}
  \mathrm{DSE}(\bm{r})
  = -\sum_{k=1}^{K} \frac{|C_k|}{N}\log\frac{|C_k|}{N}.
\end{equation}
DSE is semantic entropy with uniform sample weights; likelihood-weighted SE is
therefore not reported separately.

\paragraph{P(True).}
\textbf{P(True)}~\citep{kadavath2022language} is implemented as a black-box
cluster-agreement proxy: the fraction of samples in the same semantic cluster
as the first response
($\mathrm{P(True)}_{\mathrm{proxy}} =
|\{i: \mathrm{cl}(r_i)=\mathrm{cl}(r_1)\}|/N$);
uncertainty is $1 - \mathrm{P(True)}_{\mathrm{proxy}}$.
It is monotone in DSE here and is not treated as independent evidence.

\paragraph{SelfCheckGPT.}
\citet{manakul2023selfcheckgpt} measure the fraction of response pairs flagged
as contradictions by NLI:
\begin{equation}
  \mathrm{SCG}
  = \frac{1}{|\mathcal{P}|}
    \sum_{(i,j)\in\mathcal{P}}
    \bigl[\mathrm{NLI}(r_i,r_j)=\text{contradiction}\bigr],
\end{equation}
where $\mathcal{P}$ is the set of ordered response pairs.
For these short-answer tasks, the implementation uses whole-response NLI.

\paragraph{SRE-UQ.}
\citet{vipulanandan2026sreuq} measure perturbation sensitivity of the response
embedding distribution.
Let $\Delta_i$ denote the first-order perturbation score of response embedding
$\bm{\phi}_i$ around the weighted kernel mean embedding, using the published
bandwidth rule.
This analysis uses
$\displaystyle\mathrm{SRE\text{-}UQ}(\bm{r})
  = \frac{1}{M}\sum_{i \in \mathcal{T}} |\Delta_i|$,
where $\mathcal{T}$ indexes the $M$ highest-amplitude modes.
The estimator is imported as written; the contribution is its KG-RAG
benchmark, not a new perturbation objective.

\paragraph{VN-Entropy (KG-RAG instantiation).}
VN-Entropy measures spectral diversity of the normalised response-embedding
Gram matrix.
With $\bm{G}=\bm{V}\bm{V}^\top$ and $\bm{\rho}=\bm{G}/N$,
\begin{equation}
  \mathrm{VN\text{-}Entropy}(\bm{r}) = S(\bm{\rho})
  = -\operatorname{tr}(\bm{\rho}\log\bm{\rho})
  = -\textstyle\sum_{i=1}^{N} \lambda_i \log \lambda_i,
  \label{eq:vn_entropy}
\end{equation}
where $\{\lambda_i\}_{i=1}^N$ are the eigenvalues of $\bm{\rho}$.
$S(\bm{\rho}) = 0$ when all responses are identical and approaches $\log N$
when responses are mutually orthogonal.
This is a cosine-Gram instantiation of the VNE family introduced by Kernel
Language Entropy \citep{nikitin2024kernellanguageentropy}; the novelty lies in
the KG-RAG benchmark.


\paragraph{SD-UQ (introduced here).}
SD-UQ is a question-conditioned embedding-dispersion statistic, distinct from
\citet{qiu2024semanticdensity}'s Semantic Density.
It removes the question direction before measuring residual answer spread: if
the sampled answers point in nearly the same residual direction, dispersion is
low, the signature of a stabilised (and possibly locked-in) answer surface.
Given the unit-norm question embedding $\hat{\bm{q}}\in\mathbb{R}^d$, define
the orthogonal projector
\begin{equation}
  \bm{P}_{q}^{\perp}
  = \bm{I} - \hat{\bm{q}}\hat{\bm{q}}^\top.
\end{equation}
The projected response matrix is
\begin{equation}
  \bm{V}_{\perp} = \bm{V}\bm{P}_{q}^{\perp},
\end{equation}
where
$\bm{H} = \bm{I} - \frac{1}{N}\mathbf{1}\mathbf{1}^\top$
is the row-centering matrix.
The thin SVD of the centred projected matrix is
\begin{equation}
  \frac{\bm{H}\bm{V}_{\perp}}{\sqrt{N}}
  = \bm{U}\bm{\Sigma}\bm{W}^\top.
\end{equation}
Let $\eta_1 \geq \cdots \geq \eta_k$ be the top-$k$ singular values on
the diagonal of $\bm{\Sigma}$.
\begin{equation}
  \mathrm{SD\text{-}UQ}(\bm{r})
  \;=\;
  \exp\!\left(\frac{1}{k}\sum_{i=1}^{k} \log(\eta_i + \varepsilon)\right),
  \label{eq:sd_uq}
\end{equation}
where $k = \min(N{-}1,\, k_{\max})$ and $\varepsilon = 10^{-12}$ for
numerical stability.
At $N=5$, this is a low-sample projected-dispersion statistic, not a full
covariance estimate.
The geometric mean penalises collapse onto a single residual direction, so a
large score requires spread across several retained singular modes.
SD-UQ needs only question and answer embeddings; its attraction is operational
low cost, not mathematical ornament.

% =============================================================================
\subsection{Retrieval-State Support Diagnostics}
\label{sec:struct_metrics}

Retrieval-state diagnostics ask whether the retrieved graph contains a usable
path from the question entities to the asserted answer entity.
They matter when $p(c \mid q) \approx \delta_{c^*}$
(Section~\ref{sec:determinism}): repeated samples then see the same context,
so answer-state agreement can be high for both correct and wrong answers.
GPS is therefore a graph-support diagnostic reported in risk orientation:
lower values mean stronger graph support for the generated answer, while
higher values mean weak or absent graph support. It is not a truth label or a
logical entailment test.

Let $\mathcal{E}_q$ and $\mathcal{E}_a$ denote the sets of KG entities linked
to the question $q$ and answer $a$, respectively, filtered to
$\mathcal{E}_a \setminus \mathcal{E}_q$ to exclude trivial self-loops.
Question entities are linked by surface and fuzzy name matching.
Answer entities are linked by the same surface matcher \emph{plus} an
embedding-based soft matcher: candidate answer spans are embedded with the
retrieval encoder and linked to any entity whose name embedding has cosine
similarity at least $\tau$ to some span.
Each linked answer entity carries a link weight $w_e$: the matched cosine for
soft links and $w_e = 1$ for surface matches.

For GPS, answer entity extraction is restricted to the \emph{primary answer span}:
the first sentence of the response, truncated to 150 characters, which
prevents later explanatory text from inflating the reachable entity count.
If no usable answer entity remains after filtering and soft linking, GPS
abstains rather than returning a spurious support score.

\paragraph{Graph Path Support (GPS; retrieval-state diagnostic).}
GPS asks how strongly the answer is reachable from the question entities
within the retrieved KG, and reports the result in risk orientation as one
minus that support.
Each linked answer entity contributes distance-weighted support
$\gamma^{|L_e-\hat L(q)|}$, where $L_e$ is the shortest qualifying path
length (under the same edge-confidence and provenance filters used at
retrieval time), $\hat L(q)$ is the expected reasoning depth for the question,
and unreachable entities contribute zero:
\begin{equation}
  \mathrm{GPS}(q, a) \;=\;
  1 -
  \frac{\sum_{e \in \mathcal{E}_a} w_e\, \gamma^{|L_e-\hat L(q)|}\,
        \mathbb{1}[e\ \mathrm{reachable}]}
       {\sum_{e \in \mathcal{E}_a} w_e}.
\end{equation}
$\mathrm{GPS} = 0$ when every linked answer entity is reachable at the
expected reasoning depth (full structural support, low uncertainty);
$\mathrm{GPS} = 1$ when none are reachable within $h$ hops (no structural
support, high uncertainty); with $\gamma = 1$ the score reduces to the
unweighted unreachable-entity fraction.
The linking threshold and decay were calibrated on the RealMedQA development
run ($\tau = 0.60$, $\gamma = 0.4$) and applied frozen elsewhere; a
$5\times5$ sweep over $\tau$ and $\gamma$
(Appendix~\ref{sec:appendix_gps_sensitivity}) shows the held-out AUROCs are
stable across the grid, so the reported numbers are not an artefact of the
calibrated cell.
For 2WikiMultiHopQA and MuSiQue, $\hat L(q)$ is the logged per-question hop
count; for HotpotQA variants it is the nominal two-hop depth; for RealMedQA it
is $1$, which makes GPS identical to the calibrated one-hop-decay score on
the clinical domain.
If the system anchors to the wrong entity but traverses a locally coherent
neighbourhood, GPS can still be $0$: it scores graph support for the generated
answer, not correctness.
When no answer entities are found in the primary span (or when the
question is a direct-choice comparison), the estimator abstains and the row is
excluded from retrieval-state AUROC while being counted in the
linking-failure rate (Section~\ref{sec:grounding}).

% =============================================================================
\subsection{Evidence-State Uncertainty Measures}
\label{sec:grounding_metrics}

Evidence-state measures ask whether the retrieved passages support the answer,
so they can remain informative when sampled answers agree.
They are local answer--evidence consistency scores, not truth labels.

\paragraph{Support Entailment Uncertainty (SEU; evidence-support score).}
SEU is the normalised entailment--contradiction deficit over retrieved chunks,
close to answer-support scoring in RAGAs~\citep{es2024ragas}.

Let $c_1,\ldots,c_K$ denote the $K$ retrieved chunks and $a$ the generated answer.
Each chunk is classified by an NLI model
(\texttt{microsoft/deberta-large-mnli}; all model versions are listed in
Appendix~\ref{sec:appendix_reproducibility}):
$\ell_k \in \{E, N, C\}$ (entailment, neutral, contradiction) where
$E = \mathrm{NLI}(c_k \Rightarrow a)$.
Define the support score $s = (n_E - n_C)/K \in [-1, 1]$,
where $n_E = |\{k: \ell_k=E\}|$ and $n_C = |\{k: \ell_k=C\}|$.
\begin{equation}
  \mathrm{SEU}(q, a, \mathbf{c}) = \frac{1 - s}{2} \in [0, 1].
\end{equation}
$\mathrm{SEU} = 0.0$ when all chunks entail the answer (maximum support),
$\mathrm{SEU} = 0.5$ when chunks are all neutral (undefined support),
$\mathrm{SEU} = 1.0$ when all chunks contradict the answer (maximum conflict).
Neutral chunks therefore create a plateau at $0.5$, especially when a
general-domain NLI model declines to commit on specialist biomedical passages
(Section~\ref{sec:results_grounding}). Unlike answer-state scores, SEU needs
only one generation.

\paragraph{Faithfulness to structured provenance.}
SEU does not prove that the decoder followed the retrieved graph path.
A stricter path-faithfulness diagnostic would align answer claims to entities,
relation labels, and relation-anchor text on the retrieved paths.
The runs log those ingredients, but not sentence-level claim-to-path alignment
for every answer; path faithfulness is therefore left as a future
evidence-state diagnostic.

% =============================================================================
\subsection{Entity-Linking Quality}
\label{sec:grounding}

GPS requires matched question entities and at least one linkable answer entity.
Soft answer linking reduces abstention relative to surface matching
(Section~\ref{sec:results_structural}); yes/no answers remain a predictable
failure case.
For backward compatibility the scalar API emits a $0.5$ sentinel, but AUROC,
ECE, and precision-at-$k$ drop sentinel rows using the recorded null reason.

Entity-linking success rate $g \in [0, 1]$ measures
the fraction of content-bearing query tokens covered by matched entity names:
\begin{equation}
  g(q) = \min\!\left(1,\; \frac{|\hat{\mathcal{T}}_q^{\mathrm{cov}}|}{|\hat{\mathcal{T}}_q|}\right),
\end{equation}
where $\hat{\mathcal{T}}_q$ is the set of distinct alphanumeric content tokens
of length $\geq 4$ in $q$ (excluding stopwords), and
$\hat{\mathcal{T}}_q^{\mathrm{cov}} \subseteq \hat{\mathcal{T}}_q$ is the
subset covered by at least one matched entity name (a token $t$ is covered
when it appears as a sub-token of an entity name, or the entity name contains
$t$ as a substring).
This token-coverage score is a lightweight routing and stratification
heuristic, not an uncertainty estimator.
Appendix Table~\ref{tab:failure_modes} summarises the failure regimes of the
three families.
