% Figure: Diagnostic behaviour across retrieval regimes.
% Full-width matrix. Include with: ¥input{figures/metric_failure_modes}

¥begin{table*}[t]
¥centering
¥small
¥setlength{¥tabcolsep}{4pt}
¥renewcommand{¥arraystretch}{1.25}
¥begin{tabular}{@{}p{0.15¥linewidth}p{0.27¥linewidth}p{0.13¥linewidth}p{0.11¥linewidth}p{0.11¥linewidth}p{0.15¥linewidth}@{}}
¥toprule
¥textbf{Retrieval regime}
&
¥textbf{State of the context}
&
¥textbf{Answer-state}
&
¥textbf{SEU}
&
¥textbf{GPS}
&
¥textbf{Reading} ¥¥
¥midrule
Anchor failure
&
Entity linking fails; system falls back to text retrieval.
&
$¥sim$‾variation may persist
&
$¥sim$‾fallback passages
&
$¥oslash$‾abstains
&
Absence of a graph object is itself diagnostic. ¥¥
¥addlinespace[0.25em]
Stabilised retrieval
&
Same or near-same graph neighbourhood across samples.
&
$¥downarrow$‾compressed
&
¥checkmark‾support / contradiction
&
¥checkmark‾consistency
&
Agreement should be read against evidence and graph. ¥¥
¥addlinespace[0.25em]
Wrong coherent anchor
&
Anchoring succeeds on the wrong neighbourhood; paths locally coherent.
&
$¥downarrow$‾same wrong answer repeats
&
$¥sim$‾can warn
&
¥texttimes‾supports wrong state
&
Hardest lock-in case; trace inspection needed. ¥¥
¥addlinespace[0.25em]
KG incompleteness
&
Plausible anchor, but bridges or relations missing from the graph.
&
$¥sim$‾fallback restores variation
&
¥checkmark‾if text remains
&
$¥oslash$/$¥sim$‾reduced
&
Coverage or construction error, not decoder uncertainty. ¥¥
¥bottomrule
¥end{tabular}

¥caption{%
  ¥textbf{Expected diagnostic behaviour across KG-RAG retrieval regimes.}
  Symbols: ¥checkmark‾informative; $¥sim$‾partially informative;
  $¥downarrow$‾compressed and potentially silent; $¥oslash$‾abstains;
  ¥texttimes‾locally consistent with a wrong state.
  Each cell states what a family can observe, not a ranking;
  Table‾¥ref{tab:decomposition} defines the families and
  Table‾¥ref{tab:failure_taxonomy} gives the resulting $2{¥times}2$
  taxonomy.%
}
¥label{tab:failure_modes}
¥end{table*}
